---
title: "ecosphere_app"
output: 
  html_document: 
    keep_md: yes
date: "2024-03-07"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


## Load the Libraries
```{r}
library("tidyverse")
library("janitor")
library("naniar")
library("RColorBrewer")
```

## Load the Data
```{r}
ecosphere <- read_csv("ecosphere.csv") %>% clean_names() 
ecosphere
```

# About our App 
In our app, users will be able to select a habitat location (Grassland, Shrubland, Woodland, Wetland) and learn more about the life history events in that specific location. The life history events available for users to learn about per species and habitat include: 

Winter range size, urban tolerance, life expectancy, reproduction rates, mass(g), and diet. 


```{r}
ecosphere <- ecosphere %>%
  rename(species = scientific_name)
#Changed the old variable name of "scientific_name" to "species" to make app building easier 
```

## Top 5 Most Abundant Families 
All graphs are made using a sub-category of data with information ONLY about the 10 top most abundant families in the data. This is done to make the number of observations on the x-axis more manageable.

As such, the only families that will be included in the graphs shown in our app, are from the families listed above. 
```{r}
names(ecosphere)
```

```{r}
top5_families <- c("Anatidae", "Emberizidae", "Parulidae", "Scolopacidae", "Tyrannidae")

top5_ecosphere <- ecosphere %>%
  select(order, diet, migratory_strategy, population_size, circles, upper_95_percent_ci, 
         family, life_expectancy, log10_mass, winter_range_area, feeder_bird, common_name, 
         habitat, mean_eggs_per_clutch, range_in_cbc, median_trend, species, urban_affiliate, 
         mean_eggs_per_clutch, strata, lower_95_percent_ci) %>%
  filter(family %in% top5_families)

top5_ecosphere
```

```{r}
emberizidae <- top5_ecosphere %>% 
  filter(family=="Emberizidae")
emberizidae
```


## Making the App 

## Loading the Libraries
```{r}
library(tidyverse)
library(shiny)
#install.packages("shinydashboard")
library(shinydashboard)
```

```{r}
library(shiny)
library(shinydashboard)
library(ggplot2)
library(dplyr)

# Assuming you have your dataset stored in emberizidae
# emberizidae <- read.csv("your_dataset.csv"

# Define your dataset
emberizidae <- data.frame(
  habitat = c("Grassland", "Grassland", "Shrubland", "Shrubland", "Woodland", "Woodland", "Wetland", "Wetland"),
  diet = c("Invertebrates", "Omnivores", "Vertebrates", "Invertebrates", "Omnivores", "Vertebrates", "Invertebrates", "Omnivores"),
  life_expectancy = c("Short", "Middle", "Long", "Short", "Middle", "Long", "Short", "Middle")
)

ui <- dashboardPage(
  dashboardHeader(title = "Emberizidae"),
  dashboardSidebar(
    sidebarMenu(
      menuItem("Dashboard", tabName = "dashboard", icon = icon("pen")),
      menuItem("Ecosphere Variables", tabName = "information", icon = icon("list"))
    )
  ),
  dashboardBody(
    tabItems(
      tabItem(tabName = "dashboard",
              fluidRow(
                box(title = "Select Habitat!", width = 12,
                    selectInput("x", "Habitat Type", choices = c("Grassland", "Shrubland", "Woodland", "Wetland"),
                                selected = "Grassland"),
                    plotOutput("plot1"),
                    plotOutput("plot2")
                )
              )
      ),
      tabItem(tabName = "information",
              h2("Ecosphere Variables Section"),
              fluidRow(
                box(title = "Defining our Variables",width= 11,
                    HTML("<ul>
                          <li>Winter Range Size: The area of species’ winter range (in square kilometers)</li>
                          <li>Urban Tolerance:
                            <ul>
                              <li>No: Species is not synanthropic (DO NOT LIVE NEAR HUMANS)</li>
                              <li>Yes: Species is synanthropic (LIVE NEAR HUMANS)</li>
                            </ul>
                          </li>
                          <li>Life Expectancy: 
                            <ul>
                              <li>Short: Lifespan is <10 years in the wild</li>
                              <li>Middle: Lifespan is 10-20 years in the wild</li>
                              <li>Long: Lifespan is >20 years in the wild</li>
                            </ul>
                          </li>
                          <li>Reproduction Rates: 
                            <ul>
                              <li>mean eggs per clutch: the mean # of eggs laid per clutch</li>
                              <li>mean age of sexual maturity: mean age at which the species becomes sexually mature</li>
                              <li>Mass(g): Body mass in grams</li>
                          </ul>
                          </li>
                          <li>Diet: 
                            <ul>
                              <li>Invertebrates: primarily feed on invertebrates ex: insects</li>
                              <li>Omnivores: diet is composed of both plants and animal matter</li>
                              <li>Vertebrates: primarily feed on vertebrates by hunting or scavenging</li>
                            </ul>
                          </li>
                        </ul>")
                )  # closes box
              )
      )
    )
  )
)

server <- function(input, output, session) {
  output$plot1 <- renderPlot({
    filtered_data <- emberizidae %>%
      filter(habitat == input$x)
    
    ggplot(filtered_data, aes(x = habitat, y = diet, fill = diet)) + 
      geom_col(color = "black", alpha = 0.75, position = "dodge") +
      scale_fill_brewer(palette = "PuBuGn") +
      theme() +
      labs(title = "Diet by Habitat", 
           x = "Habitat",
           y = "Diet")
  })
  
  output$plot2 <- renderPlot({
    filtered_data2 <- emberizidae %>%
      filter(habitat == input$x)
    
    ggplot(filtered_data2, aes(x = habitat, y = life_expectancy, fill = life_expectancy)) +
      geom_bar(color = "black", alpha = 0.75, position = "dodge") +
      scale_fill_brewer(palette = "PuBuGn") +
      theme() +
      labs(title = "Life Expectancy by Habitat", 
           x = "Habitat",
           y = "Life Expectancy")
  })
  
  session$onSessionEnded(stopApp)
}

shinyApp(ui, server)



```