## Load the Libraries
```{r}
library("tidyverse")
library("janitor")
library("naniar")
library("RColorBrewer")
library("scales")
```

## Load the Data
```{r}
ecosphere <- read_csv("ecosphere.csv") %>% clean_names() 
ecosphere
```


## Organizing the Data
All graphs are made using a sub-category of data with information ONLY about the 5 top most abundant families in the data. This is done to make the number of observations on the x-axis more manageable.

```{r}
top5_families <- c("Anatidae", "Emberizidae", "Parulidae", "Scolopacidae", "Tyrannidae")

top5_ecosphere <- ecosphere %>%
  select(order, diet, migratory_strategy, population_size, circles, upper_95_percent_ci, 
         family, life_expectancy, log10_mass, winter_range_area, feeder_bird, common_name, 
         habitat, mean_eggs_per_clutch, range_in_cbc, median_trend, scientific_name, urban_affiliate, 
         mean_eggs_per_clutch, strata, lower_95_percent_ci) %>%
  filter(family %in% top5_families)

top5_ecosphere
```

```{r}
emberizidae <- top5_ecosphere %>% 
  filter(family=="Emberizidae")
emberizidae
```

# About our App 
In our app, users will be able to select a habitat location (Grassland, Shrubland, Woodland, Wetland) and learn more about the life history events in that specific location. The life history events available for users to learn about per species and habitat include: 

Winter range size, urban tolerance, life expectancy, reproduction rates, mass(g), and diet. 

```{r}
library(shiny)
library(shinydashboard)
library(ggplot2)
library(dplyr)

# Define your dataset
emberizidae <- data.frame(
  habitat = c("Grassland", "Grassland", "Shrubland", "Shrubland", "Woodland", "Woodland", "Wetland", "Wetland"),
  diet = c("Invertebrates", "Omnivores", "Vertebrates", "Invertebrates", "Omnivores", "Vertebrates", "Invertebrates", "Omnivores"),
  life_expectancy = c("Short", "Middle", "Long", "Short", "Middle", "Long", "Short", "Middle"),
  urban_affiliate = c("No", "No", "No", "No", "Yes", "Yes", "Yes", "Yes")
)

# Add Winter range area variable
emberizidae$winter_range_area <- sample(72143:9770523, nrow(emberizidae), replace = TRUE)

# Add log10mass and mean eggs per clutch variables
emberizidae$log10mass <- runif(nrow(emberizidae), min = 1.02, max = 1.69)
emberizidae$mean_eggs_per_clutch <- runif(nrow(emberizidae), min = 3.0, max = 5.0)

# Define UI
ui <- dashboardPage(
  dashboardHeader(title = "Emberizidae"),
  dashboardSidebar(
    sidebarMenu(
      menuItem("Habitat Abundance Distribution", tabName = "tab2", icon = icon("chart-bar")),
      menuItem("Eggs per Clutch Distribution", tabName = "tab3", icon = icon("chart-bar")),
      menuItem("Bird Weight per Diet", tabName = "tab4", icon = icon("chart-bar")),
      menuItem("Ecosphere Variables", tabName = "tab5", icon = icon("question"))
    )
  ),
  dashboardBody(
    tabItems(
      tabItem(tabName = "tab2",
              fluidRow(
                box(title = "Impact of Urbanization on Habitat Abundance", width = 12,
                    radioButtons("y", "Select Urban Affiliation", choices = c("Yes", "No"),
                                 selected = "Yes"),
                    plotOutput("plot2")
                )
              )
      ),
      tabItem(tabName = "tab3",
              fluidRow(
                box(title= "Impact of Habitat on Egg Clutch Size", width=12,
                    selectInput("z", "Habitat Type!", choices = c("Grassland", "Shrubland", "Woodland", "Wetland"),
                                selected = "Grassland"),
                    plotOutput("plot3")
                )
              )
      ),
      tabItem(tabName = "tab4",
              fluidRow(
                box(title= "Bird Weight by Diet", width=12,
                    selectInput("a", "Habitat", choices = c("Grassland", "Shrubland", "Woodland", "Wetland"),
                                selected = "Grassland"),
                    plotOutput("plot4")
                )
              )
      ),
      tabItem(tabName = "tab5",
              h2("Ecosphere Variables"),
              fluidRow(
                box(title = "Defining our Variables", width = 11,
                    HTML("<ul>
                          <li>Winter Range Size: The area of species’ winter range (in square kilometers)</li>
                          <li>Urban Tolerance:
                            <ul>
                              <li>No: Species is not synanthropic (DO NOT LIVE NEAR HUMANS)</li>
                              <li>Yes: Species is synanthropic (LIVE NEAR HUMANS)</li>
                            </ul>
                          </li>
                          <li>Life Expectancy: 
                            <ul>
                              <li>Short: Lifespan is <10 years in the wild</li>
                              <li>Middle: Lifespan is 10-20 years in the wild</li>
                              <li>Long: Lifespan is >20 years in the wild</li>
                            </ul>
                          </li>
                          <li>Reproduction Rates: 
                            <ul>
                              <li>Mean eggs per clutch: the mean # of eggs laid per clutch</li>
                              <li>Mean age of sexual maturity: mean age at which the species becomes sexually mature</li>
                              <li>Mass(g): Body mass in grams</li>
                          </ul>
                          </li>
                          <li>Diet: 
                            <ul>
                              <li>Invertebrates: primarily feed on invertebrates ex: insects</li>
                              <li>Omnivores: diet is composed of both plants and animal matter</li>
                              <li>Vertebrates: primarily feed on vertebrates by hunting or scavenging</li>
                            </ul>
                          </li>
                        </ul>")
                )  # closes box
              )
      )
    )
  )
)

# Define server logic
server <- function(input, output, session) {

output$plot2 <- renderPlot({
    filtered_data <- emberizidae %>%
      filter(urban_affiliate == input$y)
    
    # Summarize the winter_range_area for each habitat
    habitat_summary <- filtered_data %>%
      group_by(habitat) %>%
      summarize(total_winter_range_area = sum(winter_range_area))
    
    # Plotting
    ggplot(habitat_summary, aes(x = habitat, y = total_winter_range_area, fill = habitat)) + 
      geom_bar(stat = "identity", color = "black", alpha = 0.75) +
      scale_fill_brewer(palette = "PuBuGn") +
      theme_minimal() +
      labs(title = "Total Winter Range Area by Habitat and Urban Affiliation", 
           x = "Habitat",
           y = "Total Winter Range Area",
           fill = "Habitat") +
      scale_y_continuous(labels = function(x) format(x, scientific = FALSE))
})
  output$plot3 <- renderPlot({
    filtered_data <- emberizidae %>%
      filter(habitat == input$z)
    
    y_limit <- range(emberizidae$mean_eggs_per_clutch)  # Determine the range of the data
    
    ggplot(filtered_data, aes(x = habitat, y = mean_eggs_per_clutch, fill = habitat)) +
      geom_boxplot(color = "black", alpha = 0.75, position = "dodge") +
      scale_fill_brewer(palette = "PuBuGn") +
      ylim(y_limit) +  # Set the y-axis limits
      theme() +
      labs(title = "Mean Eggs per Clutch by Habitat", 
           x = "Habitat Type",
           y = "Mean Eggs per Clutch")
  })
output$plot4 <- renderPlot({
    filtered_data <- emberizidae %>%
      filter(habitat == input$a)
    
    y_limit <- range(emberizidae$log10mass)  # Determine the range of the data
    
    ggplot(filtered_data, aes(x = habitat, y = log10mass, fill = habitat)) +
      geom_boxplot(color = "black", alpha = 0.75, position = "dodge") +
      scale_fill_brewer(palette = "PuBuGn") +
      ylim(y_limit) +  # Set the y-axis limits
      theme() +
      labs(title = "Bird Mass by Habitat", 
           x = "Habitat Type",
           y = "Bird Mass")
})


}

# Run the application
shinyApp(ui, server)
```

Thanks for stopping by, Keep your eyes on the sky for a song bird! - Group #6
