---
title: "ecosphere_app"
output: 
  html_document: 
    keep_md: yes
date: "2024-03-07"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


## Load the Libraries
```{r}
library("tidyverse")
library("janitor")
library("naniar")
library("RColorBrewer")
```

## Load the Data
```{r}
ecosphere <- read_csv("ecosphere.csv") %>% clean_names() 
ecosphere
```

## About our App 
In our app, users will be able to select a habitat location and species to learn more about that species's life history events. The life history events available for users to learn about per species and habitat include: 

Migration/range size, urban tolerance, population size, life expectancy, reproduction rates, log10mass, and diet. 


```{r}
ecosphere <- ecosphere %>%
  rename(species = scientific_name)
#Changed the old variable name of "scientific_name" to "species" to make app building easier 
```

## Top 10 Most Abundant Families 
All graphs are made using a sub-category of data with information ONLY about the 10 top most abundant families in the data. This is done to make the number of observations on the x-axis more manageable.

```{r}
ecosphere %>% 
  select(order, family, common_name, diet, log10_mass) %>% 
  group_by(family) %>% 
  summarise(individuals_per_family = n()) %>% 
  arrange(desc(individuals_per_family)) %>%
  top_n(10) 

top10_ecosphere
```
As such, the only families that will be included in the graphs shown in our app, are from the families listed above. 
```{r}
names(ecosphere)
```


```{r}
top10_families <- c("Anatidae", "Emberizidae", "Parulidae", "Scolopacidae", 
                    "Tyrannidae", "Accipitridae", "Icteridae", "Laridae", "Corvidae")

top10_ecosphere <- ecosphere %>%
  select(order, diet, migratory_strategy, population_size, circles, upper_95_percent_ci, 
         family, life_expectancy, log10_mass, winter_range_area, feeder_bird, common_name, 
         habitat, mean_eggs_per_clutch, range_in_cbc, median_trend, species, urban_affiliate, 
         mean_eggs_per_clutch, strata, lower_95_percent_ci) %>% 
  filter(family %in% top10_families)

top10_ecosphere
```


## Creating the graphs 
Each graph will be specific to the family the user selects 

*Panchia mentioned being able to select the habitat on the app, would need to check which families exist in different habitats 

```{r}
#Bird mass and diet 
top10_ecosphere%>% 
  ggplot(aes(x=diet,y=log10_mass,fill=diet))+
  geom_boxplot(color="black", alpha=0.75)+
  scale_fill_brewer(palette ="PuBuGn")+
  theme(axis.text.x = element_text(angle = 60, hjust=1))+
  labs(title="Diet Composition and Bird Weight",
       x= "Type of Diet",
       y="Bird Weight",
       fill="Diet")+
  facet_wrap(~family)
```
```{r}
top10_ecosphere%>% 
  ggplot(aes(x=diet,y=log10_mass,fill=diet))+
  geom_boxplot(color="black", alpha=0.75)+
  scale_fill_brewer(palette ="PuBuGn")+
  theme(axis.text.x = element_text(angle = 60, hjust=1))+
  labs(title="Diet Composition and Bird Weight",
       x= "Type of Diet",
       y="Bird Weight",
       fill="Diet")
```


```{r}
#Range Size and Migration 

# Removing scientific notation
custom_format <- function(x) {
  format(x, scientific = FALSE)
}

top10_ecosphere %>% 
  ggplot(aes(x = migratory_strategy, y = winter_range_area, fill=family)) +
  geom_boxplot(color = "black", alpha = 0.75) +
  scale_fill_brewer(palette = "PuBuGn") +
  theme(axis.text.x = element_text(angle = 60, hjust = 1)) +
  scale_y_continuous(labels = custom_format) +  # Apply custom formatting to y-axis
  labs(title = "Winter Range Area by Migratory Strategy",
       x = "Migratory Strategy",
       y = "Winter Range Area")+
  facet_wrap(~family)
```


```{r}
#Population size and Life Expectancy 

top10_ecosphere
top10_ecosphere %>% 
  ggplot(aes(y = population_size, x = life_expectancy, fill=family)) +
  geom_col(color = "black", alpha = 0.75, position="dodge") +
  scale_fill_brewer(palette = "PuBuGn") +
  theme(axis.text.x = element_text(angle = 60, hjust = 1)) +
  scale_y_continuous(labels = custom_format) +  # Apply custom formatting to y-axis
  labs(title = "Population Size by Life Expectancy",
       y = "Population Size",
       x = "Life Expectancy")
```


```{r}
#Population size and Reproduction (Egg Count)
custom_format <- function(x) {
  format(x, scientific = FALSE)
}

top10_ecosphere %>% 
  ggplot(aes(x = population_size, y = mean_eggs_per_clutch, fill=family)) +
  geom_point(color = "black", alpha = 0.75, position="dodge") +
  scale_color_brewer(palette = "PuBuGn")+
  coord_flip()
```
## Creating the App 

This is just a skeleton of an app based on some experimentation with ChatGPT with built in data. Once we figure out what graphs we want to include, we just have to replace all the variable names

```{r}
library(shiny)
library(shinydashboard)
library(ggplot2)

# Define UI
ui <- dashboardPage(
  dashboardHeader(title = "Ecosphere"),
  dashboardSidebar(),  # No sidebar defined
  dashboardBody(
    titlePanel("Ecosphere Winter Birds: Species Analysis"),  # Set title for browser tab
    tabItems(
      tabItem(tabName = "species_analysis",
              h2("Species Analysis"),
              fluidRow(
                box(title = "Select Habitat",
                    radioButtons("habitat", "Habitat:",
                                 choices = unique(top10_ecosphere$habitat),
                                 selected = unique(top10_ecosphere$habitat)[1])
                ),
                box(title = "Select Species",
                    selectInput("species", "Species:",
                                choices = NULL)
                )
              ),
              fluidRow(
                plotOutput("selected_plot")
              ),
              fluidRow(
                actionButton("switch_button", "Switch Graph")
              )
      )
    )
  )
)

# Define server logic
server <- function(input, output, session) {
  observe({
    updateSelectInput(session, "species",
                      choices = setdiff(unique(top10_ecosphere$species), input$habitat))
  })
  
  selected_data <- reactive({
    if (!is.null(input$species)) {
      top10_ecosphere[top10_ecosphere$species == input$species, ]
    } else {
      top10_ecosphere[top10_ecosphere$habitat == input$habitat, ]
    }
  })
  
  plot_type <- reactiveVal("population_vs_life_expectancy")
  
  output$selected_plot <- renderPlot({
    plot_func <- switch(plot_type(),
                        "population_vs_life_expectancy" = {
                          ggplot(selected_data(), aes(x = population_size, y = life_expectancy, color = species)) +
                            geom_point() +
                            labs(x = "Population Size", y = "Life Expectancy") +
                            ggtitle("Population Size vs Life Expectancy")
                        },
                        "population_vs_mass" = {
                          ggplot(selected_data(), aes(x = population_size, y = log10_mass, color = species)) +
                            geom_point() +
                            labs(x = "Population Size", y = "Log10 Mass") +
                            ggtitle("Population Size vs Log10 Mass")
                        },
                        "life_expectancy_vs_mass" = {
                          ggplot(selected_data(), aes(x = life_expectancy, y = log10_mass, color = species)) +
                            geom_point() +
                            labs(x = "Life Expectancy", y = "Log10 Mass") +
                            ggtitle("Life Expectancy vs Log10 Mass")
                        })
    print(plot_func)
  })
  
  observeEvent(input$switch_button, {
    if (plot_type() == "population_vs_life_expectancy") {
      plot_type("population_vs_mass")
    } else if (plot_type() == "population_vs_mass") {
      plot_type("life_expectancy_vs_mass")
    } else {
      plot_type("population_vs_life_expectancy")
    }
  })
}

# Run the application
shinyApp(ui = ui, server = server)

```

