---
title: "Log10Mass_and_Diet"
output: 
  html_document: 
    keep_md: yes
date: "2024-03-07"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## R Markdown

##Background and Analysis Goals 

For our project, we are interested in analyzing the life history traits of the winter birds in the "Population trends for North American winter birds based on hierarchical models". We seek to determine which combination of life history variables results in the lowest extinction risk of winter birds in the context of increasing concern with climate change. As specified in the README file, we will only be focusing on the filter "Emberizidae" bird family.

In this rmd file, we will specifically look at `Log10mass` and `Diet` from the `ecosphere` data. We seek to determine which family of birds has the lowest extinction risk only based on their survival ability from weight and diet composition. 

## Load the Libraries

```{r}
library("tidyverse")
library("janitor")
library("naniar")
library("RColorBrewer")
```

## Load the data

```{r}
ecosphere <- read_csv("ecosphere.csv") %>% clean_names()
```

## Load the colors

```{r}
?RColorBrewer
```

```{r}
colors <- brewer.pal(9,"PuBuGn")
display.brewer.pal(9,"PuBuGn")
```

```{r}
?RColorBrewer
```

## Understanding the Data Structure

```{r}
names(ecosphere)
```

```{r}
glimpse(ecosphere)
```

As shown above, we have a combination of categorical and continous data. This will allow us to perform a variety of analysis and representation of data on various graph types.

```{r}
miss_var_summary(ecosphere)
```

```{r}
summary(ecosphere)
```

Through a combination of `miss_var_summary` and `summary` functions we can check there are no NA placeholders in the data. There does not seem to be unusual data values as indicated by the `summary` function. As such, we can assume all NA values are correclty accounted for with NA.

## Weight and Diet Analysis

```{r}
ecosphere %>% 
  select(order, family, common_name, diet, log10_mass) %>% 
  group_by(family) %>% 
  summarise(individuals_per_family=n()) %>% 
  arrange(desc(individuals_per_family))
```

As shown above, the top 10 most recorded observations of bird families. To simplify our analysis, given the extent of families listed in the data, we will only use these top 10 families of birds. 

*Plot form of top 10 bird families*

```{r}
ecosphere %>% 
  select(order, family, common_name, diet, log10_mass) %>% 
  group_by(family) %>% 
  summarise(individuals_per_family = n()) %>% 
  arrange(desc(individuals_per_family)) %>%
  top_n(5) %>%  # Filter top 10 most abundant families
  ggplot(aes(x = reorder(family, individuals_per_family), y = individuals_per_family, fill=family)) +
  geom_col(color = "black", alpha = 0.75) +
  scale_fill_brewer(palette = "PuBuGn") +
  theme() +
  labs(
    title = "Number of Individuals per Family (Top 5)",
    x = "Family",
    y = "Number of Individuals",
    fill = "Family"
  ) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```
Filtering to just look at data pertaining to the Emberizidae family. 
```{r}
library(ggplot2)

selected_colors <- c("#e0f3db", "#a8ddb5", "#7bccc4", "#4eb3d3", "#2b8cbe")

ecosphere %>%
  select(family, diet) %>%
  filter(family == "Emberizidae") %>% 
  ggplot(aes(x = family, fill = diet)) +
  geom_bar(color="black", alpha=0.75, position="dodge") +
  scale_fill_manual(values = selected_colors) +
  theme() +
  labs(
    title = "Diet Composition of Emberizidae Familiy",
    x = "Family",
    y = "Number of Birds",
    fill = "Diet"
  ) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

```{r}
ecosphere %>%
  filter(family =="Emberizidae") %>% 
  ggplot(aes(x = diet, fill = diet)) +
  geom_bar(color = "black", alpha = 0.75) +
  geom_text(stat = "count", aes(label = ..count..), vjust = -0.5) +  # Add text labels
  scale_fill_brewer(palette = "PuBuGn") +
  theme() +
  labs(
    title = "Diet Composition in Emberizidae Family",
    x = "Type of Diet",
    y = "Number of Individuals",
    fill= "Diet"
  ) 
```

We have a large abundance of Invertebrates in the data.

*Relationship between bird weight and diet* 
In the article, authors make note that bird species that are most prone to extinction events in the context of climate change are those larger in overall body weight. 

```{r}
ecosphere %>% 
  filter(family =="Emberizidae") %>% 
  ggplot(aes(x=diet,y=log10_mass,fill=diet))+
  geom_boxplot(color="black", alpha=0.75)+
  scale_fill_brewer(palette ="PuBuGn")+ 
  theme(axis.text.x = element_text(angle = 60, hjust=1))+
  labs(title="Diet Composition and Emberizidae Family Weight",
       x= "Type of Diet",
       y="Bird Weight",
       fill="Diet")
```

Determining average weight per diet. Mass is assumed to be recorded in grams, note that weight is measured in log10 scale. 
```{r}
ecosphere %>% 
  filter(family == "Emberizidae") %>% 
  group_by(diet) %>% 
  summarise(mean_weight=mean(log10_mass))
```
```{r}
ecosphere %>%
  filter(family == "Emberizidae") %>% 
  group_by(diet) %>%
  summarise(mean_weight = mean(log10_mass)) %>%
  ggplot(aes(x = diet, y = mean_weight, fill = diet)) +
  geom_col(color = "black", alpha = 0.75) +
  scale_fill_brewer(palette = "PuBuGn") +
  theme(axis.text.x = element_text(angle = 60, hjust = 1)) +
  labs(
    title = "Diet Composition and Mean Bird Weight",
    x = "Diet Composition",
    y = "Mean Bird Weight",
    fill="Diet"
  ) +
  geom_text(aes(label = round(mean_weight, 2)), vjust = -0.2, size = 3, color = "black")

```

```{r}
ecosphere %>% 
  filter(family =="Emberizidae") %>% 
  filter(habitat!= "NA") %>% 
  filter(habitat!= "Various") %>% 
  ggplot(aes(x=diet,y=log10_mass,fill=diet))+
  geom_boxplot(position="dodge", color="black", alpha=0.75)+
  scale_fill_brewer(palette ="PuBuGn")+
  theme(axis.text.x = element_text(angle = 60, hjust=1))+
  labs(title="Diet Composition and Bird Weight",
       x= "Diet Composition",
       y="Bird Weight",
       fill="Diet")+
  facet_wrap(~habitat)
```

In the graoh above, we see that while members of the Emberizidae family have the capability to exist in variety of habitats (exluding the ocean), there is a correlation of what habitats they do best in based on the range of body weights observed. We observe see diets in grasslands, shurblands, and woodlands,however it is critical to note the shifting mean of body weight birds in these different habitats are showing. Meaning, a seed diet can be better or worse depending on the habitat. A similar pattern is obsereved in the omnivore and invertebrate diets as well. 

