---
title: "Log10Mass_and_Diet"
output: 
  html_document: 
    keep_md: yes
date: "2024-03-07"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## R Markdown

##Background and Analysis Goals 

For our project, we are interested in analyzing the life history traits of the winter birds in the "Population trends for North American winter birds based on hierarchical models". We seek to determine which combination of life history variables results in the lowest extinction risk of winter birds in the context of increasing concern with climate change. 

In this rmd file, we will specifically look at `Log10mass` and `Diet` from the `ecosphere` data. We seek to determine which family of birds has the lowest extinction risk only based on their survival ability from weight and diet composition. 

## Load the Libraries

```{r}
library("tidyverse")
library("janitor")
library("naniar")
library("RColorBrewer")
```

## Load the data

```{r}
ecosphere <- read_csv("ecosphere.csv") %>% clean_names()
```

## Load the colors

```{r}
?RColorBrewer
```

```{r}
colors <- brewer.pal(9,"PuBuGn")
display.brewer.pal(9,"PuBuGn")
```

```{r}
?RColorBrewer
```

## Understanding the Data Structure

```{r}
names(ecosphere)
```

```{r}
glimpse(ecosphere)
```

As shown above, we have a combination of categorical and continous data. This will allow us to perform a variety of analysis and representation of data on various graph types.

```{r}
miss_var_summary(ecosphere)
```

```{r}
summary(ecosphere)
```

Through a combination of `miss_var_summary` and `summary` functions we can check there are no NA placeholders in the data. There does not seem to be unusual data values as indicated by the `summary` function. As such, we can assume all NA values are correclty accounted for with NA.

## Weight and Diet Analysis

```{r}
ecosphere %>% 
  select(order, family, common_name, diet, log10_mass) %>% 
  group_by(family) %>% 
  summarise(individuals_per_family=n()) %>% 
  arrange(desc(individuals_per_family))
```

As shown above, the top 10 most recorded observations of bird families. To simplify our analysis, given the extent of families listed in the data, we will only use these top 10 families of birds. 

*Plot form of top 10 bird families*

```{r}
ecosphere %>% 
  select(order, family, common_name, diet, log10_mass) %>% 
  group_by(family) %>% 
  summarise(individuals_per_family = n()) %>% 
  arrange(desc(individuals_per_family)) %>%
  top_n(10) %>%  # Filter top 10 most abundant families
  ggplot(aes(x = reorder(family, individuals_per_family), y = individuals_per_family, fill=family)) +
  geom_col(color = "black", alpha = 0.75) +
  scale_fill_brewer(palette = "PuBuGn") +
  theme() +
  labs(
    title = "Number of Individuals per Family (Top 10)",
    x = "Family",
    y = "Number of Individuals",
    fill = "Family"
  ) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```
```{r}
library(ggplot2)

selected_colors <- c("#e0f3db", "#a8ddb5", "#7bccc4", "#4eb3d3", "#2b8cbe")

ecosphere %>%
  select(family, diet) %>%
  filter(family %in% c("Anatidae", "Emberizidae", "Parulidae", "Scolopacidae", "Picidae", "Tyrannidae", "Accipitridae", "Icteridae", "Laridae", "Corvidae")) %>% 
  ggplot(aes(x = family, fill = diet)) +
  geom_bar(color="black", alpha=0.75) +
  scale_fill_manual(values = selected_colors) +
  theme() +
  labs(
    title = "Diet Composition of Top 10 Bird Families",
    x = "Family",
    y = "Number of Birds",
    fill = "Diet"
  ) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```



```{r}
table(ecosphere$diet)
```

```{r}
ecosphere %>%
  filter(family %in% c("Anatidae", "Emberizidae", "Parulidae", "Scolopacidae", "Picidae", "Tyrannidae", "Accipitridae", "Icteridae", "Laridae", "Corvidae")) %>% 
  ggplot(aes(x = diet, fill = diet)) +
  geom_bar(color = "black", alpha = 0.75) +
  geom_text(stat = "count", aes(label = ..count..), vjust = -0.5) +  # Add text labels
  scale_fill_brewer(palette = "PuBuGn") +
  theme() +
  labs(
    title = "Diet Composition in Top 10 Recorded Winter Birds",
    x = "Type of Diet",
    y = "Number of Individuals",
    fill= "Diet"
  ) 
```

We have a large abundance of Invertebrates in the data.

*Relationship between bird weight and diet* 
In the article, authors make note that bird species that are most prone to extinction events in the context of climate change are those larger in overall body weight. 

```{r}
ecosphere %>% 
  filter(family %in% c("Anatidae", "Emberizidae", "Parulidae", "Scolopacidae", "Picidae", "Tyrannidae", "Accipitridae", "Icteridae", "Laridae", "Corvidae")) %>% 
  ggplot(aes(x=diet,y=log10_mass,fill=diet))+
  geom_boxplot(color="black", alpha=0.75)+
  scale_fill_brewer(palette ="PuBuGn")+ #why is color not working
  theme(axis.text.x = element_text(angle = 60, hjust=1))+
  labs(title="Diet Composition and Bird Weight",
       x= "Type of Diet",
       y="Bird Weight",
       fill="Diet")
```

This graph makes it clear that winter birds who rely on nectar as a food source, have significantly smaller weights. While this does not mean this winter bird is doing poor, it demonstrates a difference in survival tactics depending on the diet. In the context of climate change, flowering plants are expected to do poor in habitats with rising temperatures. While more flowers are found to bloom with warmer temperatures, the amount of nectar found is not increasing proportionally. Thus suggesting a decrease in food, which is bound to have negative impacts for winter birds that depend on nectar for survival. From this fact, nectar dependent birds will be excluded from being bird species `less` vulnerable to extinction on the basis of weight.

Determining average weight per diet. Mass is assumed to be recorded in grams, note that weight is measured in log10 scale. 
```{r}
ecosphere %>% 
  group_by(diet) %>% 
  summarise(mean_weight=mean(log10_mass))
```
```{r}
ecosphere %>%
  filter(family %in% c("Anatidae", "Emberizidae", "Parulidae", "Scolopacidae", "Picidae", "Tyrannidae", "Accipitridae", "Icteridae", "Laridae", "Corvidae")) %>% 
  group_by(diet) %>%
  summarise(mean_weight = mean(log10_mass)) %>%
  ggplot(aes(x = diet, y = mean_weight, fill = diet)) +
  geom_col(color = "black", alpha = 0.75) +
  scale_fill_brewer(palette = "PuBuGn") +
  theme(axis.text.x = element_text(angle = 60, hjust = 1)) +
  labs(
    title = "Diet Composition and Mean Bird Weight",
    x = "Diet Composition",
    y = "Mean Bird Weight",
    fill="Diet"
  ) +
  geom_text(aes(label = round(mean_weight, 2)), vjust = -0.2, size = 3, color = "black")

```


```{r}
ecosphere %>% 
  filter(habitat!= "NA") %>% 
  ggplot(aes(x=diet,y=log10_mass,fill=diet))+
  geom_boxplot(position="dodge", color="black", alpha=0.75)+
  scale_fill_brewer(palette ="PuBuGn")+
  theme(axis.text.x = element_text(angle = 60, hjust=1))+
  labs(title="Diet Composition and Bird Weight",
       x= "Diet Composition",
       y="Bird Weight",
       fill="Diet")+
  facet_wrap(~habitat)
```

From the prior graph, we learn that birds with the highest body weight are those reliant on vegation and invertebrates have the highest log10mass. 

```{r}
ecosphere %>% 
  select(order, family, common_name, diet, log10_mass) %>%
  filter(diet== "Vegetation" | diet=="Vertebrates" | diet=="Nectar") %>% 
  ggplot(aes(x=order,fill=diet))+
  geom_bar(position="dodge",color="black", alpha=0.75)+
  scale_fill_brewer(palette ="PuBuGn")+
  theme(axis.text.x = element_text(angle = 60, hjust=1))+
  facet_wrap(~diet)
```

In the study, authors conclude that body size can influence the survival status of birds, with larger bodied taxa being at higher risk. In terms of diet, taxa that occupy higher tropic levels face elevated risk of extinction. 

